Supplementary material to ``Estimating the Attack Ratio of Dengue Epidemics under Time-varying Force of Infection using Aggregated Notification Data''

### 2015-08-03 scirep_version.tex
updated supplSR content

Dengue is an arthropod-borne febrile disease caused by a flavivirus with four 
serotypes which causes an estimated  $50$ million infections each 
year~\cite{Guzman2010}.
In humans, immunity against  a particular serotype is  considered permanent 
after the exposure and cross-immunity to the other serotypes is 
generally considered short lived~\cite{halstead_dengue_2007} although some 
studies argue for a longer duration of cross-immunity \cite{Reich2013, 
salje2012}.
As a consequence, the proportions of viral serotypes co-circulating at any 
point in time are strongly dependent on previous incidence patterns of the 
disease, which determine the number of individuals susceptible to each serotype 
at any point in time.
% LM: Maybe a citation here?

Dengue transmission is also modulated by environmental conditions, among which, 
temperature, due to its effects on the 
vector reproduction, stands out as a strong predictor of 
incidence~\cite{honorio_temporal_2009,wu_higher_2009}.
In places with sufficient seasonal temperature variation, dengue is 
predominantly a  summer disease.
So it is fair to say that these environmental 
fluctuations play a key role in determining beginning and end of epidemic 
periods.
This climatic influence is exerted mainly through its effects on the force of 
infection, which cannot be taken as constant~\cite{reiner_time-varying_2014} 
but rather as a seasonal (oscillating) function of time.
The long term dynamics of dengue is also modulated by the alternation of virus 
types in circulation. Demographics also plays a role in replenishing the 
population of susceptibles.

@@

The attack ratio is also influenced by the reproductive number of the disease 
~\cite{bacaer_final_2009, katriel_attack_2012}, which is closely associated 
with the force of infection.
Thus the incorporation of the effective reproduction number, \rr, as a function 
of time, is crucial to an accurate estimation of the AR of seasonal diseases 
like dengue and Influenza.

Other methods for estimating the  number of susceptibles while 
accommodating time-varying force of infection have been proposed 
before, for measles~\cite{bjornstad_dynamics_2002, 
wallinga_reconstruction_2003}, a disease that shows remarkable seasonality.
These methods try to reconstruct the entire 
series of infectious and susceptibles from case data using deterministic models 
and generally work well for measles because there is a one-to-one 
relationship between exposure and immunity, since measles is caused by a 
single-strain pathogen.
Recently, methods in the same fashion were developed for dengue when 
serotype-specific data is available~\cite{Reiner2014}.
When such data is not available, the series of susceptibles to all possible
serotypes, cannot be reconstructed based solely on a deterministic transmission 
model, since the arrival/re-emergence of  new serotypes, an intrinsically 
stochastic events, can drastically change the pool of susceptibles, throwing off 
any sequential estimation based on the  incidence dynamics.

In this paper, we propose a new approach to estimate the number (fraction) of 
susceptibles using a simplified model of dengue transmission based on a 
single-strain Susceptible-Infectious-Removed (SIR) model with time-varying 
infection rate.
In order to bypass the limitations of not knowing the serotype-specific 
seroprevalence and the exact behaviour of the force of infection through time, 
we
propose to inform the time-varying transmissibility using the \rr series 
derived from the notification data~\cite{nishiura}.
We extend a Bayesian framework previously used to estimate the number 
of susceptibles in Influenza epidemics in Europe~\cite{pone2011} to include 
time-varying force of infection and derive a probability distribution  for 
\rr to accommodate uncertainty in the estimates. 
Then, from the incidence series and the population at risk, we  calculate the 
attack ratio for each epidemic.
We apply our method to estimate $S_0$ before every major dengue epidemic in the 
city of Rio de Janeiro, Brazil in the last 18 years.

@@

Let $Y_t$ be the number of reported disease cases for a particular time $t \in 
(0, T)$.
Nishiura el al. (2010)~\cite{nishiura} extend the theory developed by 
Stallybrass et al. (1931)~\cite{stallybrass} and propose to estimate \rr as
\begin{equation}
\label{eq:Rtestimate}
R_t = \left( \frac{Y_{t+1}}{Y_t}\right)^{1/n}
\end{equation}
where $n$ is taken to be the ratio between the length of reporting interval and 
the mean generation time of the disease.
Here we are interested in the simpler case $n=1$.
If \rr is to be used as a decision tool, however, one needs to be able to 
quantify the 
uncertainty about estimate in equation~\ref{eq:Rtestimate}. 
Here we detail how to obtain credibility intervals for \rr under the assumption 
that the counts $Y_t$ are Poisson distributed for all $t$.

We explore the approach of Ederer and Mantel~\cite{mantel}, whose objective is 
to obtain 
confidence intervals for the ratio of two Poisson counts. 
Let $Y_{t} \sim Poisson(\lambda_t)$ and $Y_{t+1} \sim Poisson(\lambda_{t+1})$ 
and define $S = Y_{t} + Y_{t+1}$.
The authors note that by conditioning on the sum $S$
\begin{align}
\label{eq:binlike}
Y_{t+1} | S &\sim Binomial(S, \theta_t) \\
\theta_t &= \frac{\lambda_{t+1}}{\lambda_{t} + \lambda_{t+1}}
\end{align}
Let $c_{\alpha}(\theta_t) = \{\theta_t^{(L)} , \theta_t^{(U)} \}$ be such that 
$Pr(\theta_t^{(L)}<\theta_t <\theta_t^{(U)}) = \alpha$.
Analogously, define $c_{\alpha}(R_t) = \{R_t^{(L)} , R_t^{(U)} \}$ such that 
$Pr(R_t^{(L)}<R_t<R_t^{(U)}) = \alpha$.
Ederer and Mantel (1974)~\cite{mantel} show that one can construct a $100\alpha 
\%$ confidence interval for \rr by noting that
\begin{equation}
\label{eq:confRt}
 R_t^{(L)} = \frac{\theta_t^{(L)}}{(1-\theta_t^{(L)})} \quad \text{and} \quad 
R_t^{(U)} = \frac{\theta_t^{(U)}}{(1-\theta_t^{(U)})}\\
\end{equation}
Because the transform from $\theta$ to \rr is monotonically 
increasing, the result holds for confidence and credibility intervals alike.

Many authors have chosen to quantify the uncertainty about $\theta$ 
following orthodox approaches  (see for example~\cite{wilson} 
and~\cite{clopper}) mainly for simplicity.
We choose instead to take a Bayesian approach and use the  $100\alpha \%$ 
posterior credibility interval for $\theta_t$ as $c_{\alpha}(\theta_t)$.
If we choose the conjugate beta prior with parameters $a_0$ and $b_0$ for the 
binomial likelihood in (\ref{eq:binlike}), the posterior distribution for 
$\theta_t$ is
\begin{equation}
\label{eq:thetapost}
p(\theta_t| Y_{t+1}, S) \sim Beta(Y_{t+1} + a_0, Y_t + b_0)
\end{equation}
Combining equations~(\ref{eq:confRt}) and~(\ref{eq:thetapost}) 
tells us that the induced posterior distribution of $R_t$ is 
a beta prime (or inverted beta) with parameters $ a_1 = Y_{t+1} + a_0$ and $b_1 
=  Y_t + b_0$~\cite{dubey1970}.
The density of the induced distribution is then 
\begin{equation}
\label{eq:densityMantel}
f_P(R_t| a_1, b_1) = \frac{\Gamma(a_1 + b_1)}{\Gamma(a_1)\Gamma(b_1)} R_t^{a_1 
- 
1} (1 + R_t)^{-(a_1 + b_1)}
\end{equation}
Thus, the expectation of \rr is $a_1/(b_1 - 1)$ and its variance is 
$a_1(a_1 + b_1 - 1)/\left((b_1 - 2)(b_1 - 1)^2 \right) $.
Note that this result holds only for $n = 1$.
Sampling from the posterior in~(\ref{eq:densityMantel}) can be made 
straightforward by first sampling from~(\ref{eq:thetapost}) and then applying 
the transform in~(\ref{eq:confRt}).
Also, one can choose $a_0$ and $b_0$ so as to elicit meaningful prior 
distributions for $R_t$.
We show how to elicit the prior for $R_t$ from specified prior mean and variance
or coefficient of variation in the Appendix.

@@

For our analysis we chose to let the force of infection vary with time, just 
as it does in the actual epidemics, as seen in the data. So as the epidemic 
progresses, the effective transmission  rate changes and is 
given by 
\begin{equation} 
 \label{eq:effbeta}
 \beta(t) = \frac{R_t\cdot\tau}{S_0}
\end{equation}
where $R_t$ is the effective reproductive number, estimated as 
in~\ref{eq:Rtestimate}.
The complete model with the time-varying force of infection is given by
the system of ordinary differential equations:
\begin{align}
   \label{eq:model}
 \frac{dS}{dt} &= -\beta(t)SI \\     \nonumber
 \frac{dI}{dt} &= \beta(t)SI - \tau I&\\      \nonumber
 \frac{dR}{dt} &= \tau I&
\end{align}  
where $S + I + R = 1 \: \forall\: t$. % \in [T_0, T_1]$. 
Of course, this is a rather simplified model, in which, for instance, the 
vector is omitted.
The rationale for this simplification is based on the ability of the 
empirically derived $R_t$  to incorporate the effects of the fluctuating vector 
populations.
Also, although there are multiple circulating serotypes, our approach
can not discriminate between them due to the lack of serotype-specific data.
Nevertheless, this modelling strategy can still provide some insight into the 
disease dynamics and allows us to estimate the initial fraction of susceptibles 
$S_0$, a key epidemiological parameter.

@@

We take a Bayesian approach to the estimation of $S_0$.
First the incidence time series was divided into $J=13$ 
epidemic windows that corresponded to significant raises in incidence and 
normalized to lie on the $[0,1]$ interval.
For a given interval $j = \{ t_j^{\text{start}}, t_j^{\text{end}} \} $ we 
observe an incidence time series $\mathbf{Y_{j}}$.
We are thus interested in the posterior distribution
\begin{equation}
 \label{eq:S0post}
 p(S_{0j}|\mathbf{Y_{j}}) \propto l(\mathbf{Y_{j}}|S_{0j}, R_t, m, \tau 
)\pi(S_{0j}) 
\end{equation}
The likelihood $l(\mathbf{Y_{j}}|\cdot)$ is assumed to be a Normal distribution 
with fixed variance $\sigma^2$.
In this estimation procedure we kept $R_t$ fixed at fixed at the posterior mean 
obtained as described above and fixed $\tau = 1/7\: \text{days}^{-1}$.
To complete the inference, we need to specify prior distributions for the 
parameters of interest.
We place a flat $\text{Beta}(1, 1)$ prior on $S_{0j}\:\forall j$

To approximate the posterior in~(\ref{eq:S0post}) we use Markov chain Monte 
Carlo techniques implemented in the Bayesian inference with Python 
(BIP)~\cite{pone2011} available at  
\url{http://code.google.com/p/bayesian-inference/}.
BIP uses a Differential Evolution Adaptive Metropolis (DREAM)~\cite{vrugt2008} 
scheme that efficiently samples from high-dimensional joint distributions using 
multiple adaptive chains running in parallel with delayed rejection.
Also, as the numerical integration routine implemented within BIP needs 
$\beta(t)$ to be available at 
arbitrary values of $t$, i.e., as continuous function of time because of the 
variable step size, we used linear interpolation to obtain values of $R_t$ for 
any time point.
In this study we used one chain per parameter, i.e, 3 chains for each run.
The chains were run until $5000$ samples were obtained after discarding $500$ 
burn-in samples.
Convergence of the parallel chains was verified at every 100 iterations by the 
calculation of the Gelman-Rubin's R (potential scale 
reduction factor), which approaches $1$ at convergence~\cite{brooks1998}.

@@

In this paper we propose a method to bypass the lack of serotype-specific case 
data by informing the time-varying force of infection with the instantaneous 
reproductive number, \rr which we calculate from aggregated data.
The main contribution of this paper can be summarized in the following items: 
(i) we show a method to quantify uncertainty 
about \rr that is readily applicable to other diseases and; (ii) we propose to 
use \rr to inform a dynamic epidemic model with time-varying force of infection 
in order to gain insight into the attack ratio of each epidemic; (iii) We 
propose an estimation procedure for circulating serotype's $S_0$ from 
aggregate case data, which is robust to epidemic sizes; (iv) We estimate the AR 
for 18 years of Dengue epidemics.

Figure~\ref{fig:rtseries}  shows the $R_t$ series, according to 
(\ref{eq:Rtestimate})~\cite{nishiura} along with the confidence bands derived 
in this paper. 
It can be seen that the inter-epidemic periods are characterized by $R_t$ being 
indistinguishable from 1.
Due to the intrinsic variability of the \rr series, the examination of its 
credible intervals is essential to identify periods of sustained transmission.
The wider intervals between epidemics are due to the scarcity of cases during 
these periods.
The method to quantify uncertainty proposed here provides more conservative 
credibility intervals, and therefore offers protection against false alarms.
\begin{center}
\begin{figure}[!h]
 \centering
 \includegraphics[width=16cm]{./plots/rt_series.png}
 \caption{{\bf Estimated time-series for $R_t$, along with 95\% credible 
intervals.} Top panel show reported cases from which $R_t$ is estimated.
As expected, uncertainty about \rr is greater when the case counts are low, for 
instance in the period $2003$--$2006$, which represented a big hiatus between 
major epidemics.
The intrinsic variability of \rr can be used to inform the time-varying force 
of infection, since it reflects variation in the vector population and other 
environmental factors such as temperature and seasonal variation.}
\label{fig:rtseries}
\end{figure}
\end{center}

A key epidemiological quantity is the attack ratio (AR) of an epidemic, a 
measure of morbidity and speed of spread which can be used to predict epidemic 
size and help efficient Public Health planning.
The AR depends fundamentally on the population at risk, which in the case of 
dengue is every naive (to a particular serotype) individual in the population.
Estimating the initial susceptible fraction $S_0$ for each epidemic is thus 
central to the estimation of the AR.
Methods for estimating the  number of susceptibles have been proposed 
before, for other diseases~\cite{bjornstad_dynamics_2002, 
wallinga_reconstruction_2003}.
These methods try to reconstruct the entire 
series of infectious and susceptibles for measles 
outbreaks from case data.
In the case of dengue, the full (multi-year/multi-epidemic)
series of susceptibles to all possible serotypes, cannot be reconstructed based 
solely on a deterministic transmission model, since the arrival/re-emergence of 
new serotypes (which are a stochastic events) can change drastically the pool 
of susceptibles throwing off any sequential estimation based on the incidence
dynamics.

Since there is very limited information regarding the actual proportions of 
each virus in circulation and most information available is about the 
predominant serotypes for some epidemics in the period of 
study only~\cite{macedo_virological_2013}, we propose the use of a 
simplified a single strain model.
The main argument we put forward is that by conditioning on the \rr series, we 
implicitly take into account the variability introduced by the 
co-circulation of multiple serotypes and heterogeneous levels of immunity in 
the general population.
We sought to deal with all important sources of uncertainty impinging on the 
estimation of the AR of a dengue epidemic, but not all could be satisfactorily 
addressed in this analysis.
For instance, in any given epidemic there is a large number of 
mild and asymptomatic cases, which nevertheless acquire immunity.
It is estimated that for every case reported, up to 10-20 are not seen by health 
authorities~\cite{luz_disability_2009}.
Another source of uncertainty is under-reporting of diagnosed cases, which is a 
serious issue in the health care systems of many developing countries such as 
Brazil. 
Duarte and Fran\c{c}a (2006)~\cite{duarte_data_2006}, estimated 
the sensitivity of Dengue reporting for hospitalized patients in 
Belo-Horizonte, Brazil to be of 63\%, meaning that approximately 37\% of the 
suspected Dengue cases go unreported.  
Lastly, demography and migrations affect the number of susceptible in ways 
which are not easy to fully determine.

@@

\begin{center}
\begin{figure}
\includegraphics[width=\textwidth]{./plots/figure-2-coelho.jpg}

  \caption{{\bf Maps showing the incidence of dengue in the city of Rio de 
Janeiro from 2010 to 2013(left-to-right, rowise).} Circles indicate individual 
notified cases. A 
heatmap is overlayed on the maps showing absolute density of cases.
It can be seen that several areas of the city were affected and no region seems 
to be free of transmission risk.
This suggests that although transmission risk varies spatially, there is 
significant exposure over the entire city. Maps were made with QGis GIS 
software(\url{qgis.org})}
  \label{fig:mapas}
\end{figure}
\end{center}

Table~\ref{tab:AR} contains the attack ratios and medians of the $S_0$ 
estimated for each epidemic/outbreak.
It is interesting to notice that the larger epidemics, in terms of peak size are 
not the one with the greater attack ratios.
This stresses the importance of knowing the immunological structure of the 
population. Knowing the $S_0$ for the circulating viruses we can order to more 
accurately assess the potential impact of a coming epidemic, since particularly 
virulent types, can be rendered less of a threat by a low $S_0$.

>>>>

Dengue is an arthropod-borne febrile disease caused by a flavivirus with four 
serotypes (DEN-1, 2, 3 and 4) which causes an estimated  $50$ million 
infections each 
year~\citep{Guzman2010}.
In humans, immunity against a particular serotype is considered permanent 
after the exposure and cross immunity to other serotypes is considered short 
lived~\citep{halstead_dengue_2007} although some 
studies argue for a longer duration of cross-immunity \citep{Reich2013, 
salje2012}.
As a consequence, the proportions of viral serotypes co-circulating at any 
point in time are strongly dependent on previous incidence patterns of the 
disease, which determine the number of individuals susceptible to each serotype 
at any point in time.
% LM: Maybe a citation here?

Dengue transmission is also modulated by environmental conditions, among which, 
temperature, due to its effects on the 
vector reproduction, stands out as a strong predictor of 
incidence~\citep{honorio_temporal_2009,wu_higher_2009}.
In places with sufficient seasonal temperature variation, dengue is 
predominantly a  summer disease.
So it is fair to say that these environmental 
fluctuations play a key role in determining beginning and end of epidemic 
periods.
This climatic influence is exerted mainly through its effects on the force of 
infection, which cannot be taken as constant~\citep{reiner_time-varying_2014} 
but rather as a seasonal (oscillating) function of time.
The long term dynamics of dengue is also modulated by the alternation of virus 
types in circulation. Demographics also plays a role in replenishing the 
population of susceptibles.

@@

The attack ratio is also influenced by the reproductive number of the disease 
~\citep{bacaer_final_2009, katriel_attack_2012}, which is closely associated 
with the force of infection.
Thus the incorporation of the effective reproduction number, \rr, as a function 
of time, is crucial to an accurate estimation of the AR of seasonal diseases 
like dengue and Influenza.

Other methods for estimating the  number of susceptibles while 
accommodating time-varying force of infection have been proposed 
before, for measles~\citep{bjornstad_dynamics_2002, 
wallinga_reconstruction_2003}, a disease that shows remarkable seasonality.
These methods try to reconstruct the entire 
series of infectious and susceptibles from case data using deterministic models 
and generally work well for measles because there is a one-to-one 
relationship between exposure and immunity, since measles is caused by a 
single-strain pathogen.
Recently, methods in the same fashion were developed for dengue when 
serotype-specific data is available~\citep{Reiner2014}.
When such data is not available, the series of susceptibles to all possible
serotypes, cannot be reconstructed based solely on a deterministic transmission 
model, since the arrival/re-emergence of  new serotypes, an intrinsically 
stochastic event, can drastically change the pool of susceptibles, throwing off 
any sequential estimation based on the  incidence dynamics.

Dengue has been reintroduced in Rio de Janeiro in 1986 after being absent for 
68 years~\citep{nogueira1999dengue}.
During the last decades of the 20th century only DEN-1 and DEN-2 serotypes were 
in circulation. The remaining serotypes DEN-3 and DEN-4, arrived respectively 
in 2000 and 2011 \citep{de2004dengue, nogueira2011dengue}. Due to this patterns 
of recent re-introduction of the disease, its incidence dynamics is still 
dominated by the introduction events and environmental determinants of 
transmission.

In this paper, we propose a new approach to estimate the number (fraction) of 
susceptibles using a simplified model of dengue transmission based on a 
single-strain Susceptible-Infectious-Removed (SIR) model with time-varying 
infection rate.
In order to bypass the limitations of not knowing the serotype-specific 
seroprevalence and the exact behaviour of the force of infection through time, 
we propose to inform the time-varying transmissibility using the \rr series 
derived from the notification data~\citep{nishiura}.
We extend a Bayesian framework previously used to estimate the number 
of susceptibles in Influenza epidemics in Europe~\citep{pone2011} to include 
time-varying force of infection and derive a probability distribution  for 
\rr to accommodate uncertainty in the estimates. 
Then, from the incidence series and the population at risk, we  calculate the 
attack ratio for each epidemic.
We apply our method to estimate $S_0$ before every major dengue epidemic in the 
city of Rio de Janeiro, Brazil in the last 18 years.

@@

Let $Y_t$ be the number of reported disease cases for a particular time $t \in 
(0, T)$.
Nishiura el al. (2010)~\citep{nishiura} extend the theory developed by 
Stallybrass et al. (1931)~\citep{stallybrass} and propose to estimate \rr as
\begin{equation}
\label{eq:Rtestimate}
R_t = \left( \frac{Y_{t+1}}{Y_t}\right)^{1/n}
\end{equation}
where $n$ is taken to be the ratio between the length of reporting interval and 
the mean generation time of the disease.
Here we are interested in the simpler case $n=1$.
If \rr is to be used as a decision tool, however, one needs to be able to 
quantify the 
uncertainty about estimate in equation~\ref{eq:Rtestimate}. 
Here we detail how to obtain credibility intervals for \rr under the assumption 
that the counts $Y_t$ are Poisson distributed for all $t$.

We explore the approach of Ederer and Mantel~\citep{mantel}, whose objective 
was  to obtain confidence intervals for the ratio of two Poisson counts. 
Let $Y_{t} \sim Poisson(\lambda_t)$ and $Y_{t+1} \sim Poisson(\lambda_{t+1})$ 
and define $W = Y_{t} + Y_{t+1}$.
The authors note that by conditioning on the sum $W$
\begin{align}
\label{eq:binlike}
Y_{t+1} | S &\sim Binomial(W, \theta_t) \\
\theta_t &= \frac{\lambda_{t+1}}{\lambda_{t} + \lambda_{t+1}}
\end{align}
Let $c_{\alpha}(\theta_t) = \{\theta_t^{(L)} , \theta_t^{(U)} \}$ be such that 
$Pr(\theta_t^{(L)}<\theta_t <\theta_t^{(U)}) = \alpha$.
Analogously, define $c_{\alpha}(R_t) = \{R_t^{(L)} , R_t^{(U)} \}$ such that 
$Pr(R_t^{(L)}<R_t<R_t^{(U)}) = \alpha$.
Ederer and Mantel (1974)~\citep{mantel} show that one can construct a 
$100\alpha 
\%$ confidence interval for \rr by noting that
\begin{equation}
\label{eq:confRt}
 R_t^{(L)} = \frac{\theta_t^{(L)}}{(1-\theta_t^{(L)})} \quad \text{and} \quad 
R_t^{(U)} = \frac{\theta_t^{(U)}}{(1-\theta_t^{(U)})}\\
\end{equation}
Because the transform from $\theta$ to \rr is monotonically 
increasing, the result holds for confidence and credibility intervals alike.

Many authors have chosen to quantify the uncertainty about $\theta$ 
following orthodox approaches  (see for example~\citep{wilson} 
and~\citep{clopper}) mainly for simplicity.
We choose instead to take a Bayesian approach and use the  $100\alpha \%$ 
posterior credibility interval for $\theta_t$ as $c_{\alpha}(\theta_t)$.
If we choose the conjugate beta prior with parameters $a_0$ and $b_0$ for the 
binomial likelihood in (\ref{eq:binlike}), the posterior distribution for 
$\theta_t$ is
\begin{equation}
\label{eq:thetapost}
p(\theta_t| Y_{t+1}, W) \sim Beta(Y_{t+1} + a_0, Y_t + b_0)
\end{equation}
Combining equations~(\ref{eq:confRt}) and~(\ref{eq:thetapost}) 
tells us that the induced posterior distribution of $R_t$ is 
a beta prime (or inverted beta) with parameters $a_1 = Y_{t+1} + a_0$ and $b_1 
=  Y_t + b_0$~\citep{dubey1970}.
The density of the induced distribution is then 
\begin{equation}
\label{eq:densityMantel}
f_P(R_t| a_1, b_1) = \frac{\Gamma(a_1 + b_1)}{\Gamma(a_1)\Gamma(b_1)} R_t^{a_1 
- 
1} (1 + R_t)^{-(a_1 + b_1)}
\end{equation}
Thus, the expectation of \rr is $a_1/(b_1 - 1)$ and its variance is 
$a_1(a_1 + b_1 - 1)/\left((b_1 - 2)(b_1 - 1)^2 \right) $.
Note that this result holds only for $n = 1$.
Sampling from the posterior in~(\ref{eq:densityMantel}) can be made 
straightforward by first sampling from~(\ref{eq:thetapost}) and then applying 
the transform in~(\ref{eq:confRt}).
Also, one can choose $a_0$ and $b_0$ so as to elicit meaningful prior 
distributions for $R_t$.
We show how to elicit the prior for $R_t$ from specified prior mean and variance
or coefficient of variation in the~\textbf{Appendix}.

@@

For our analysis we chose to let the force of infection vary with time, just 
as it does in the actual epidemics, as seen in the data. So as the epidemic 
progresses, the effective transmission  rate changes and is 
given by 
\begin{equation} 
 \label{eq:effbeta}
 \beta(t) = \frac{R_t\cdot\tau}{S_0}
\end{equation}
where $R_t$ is the effective reproductive number, estimated as 
in~\ref{eq:Rtestimate} and $S_0$ is the initial fraction of susceptible 
individuals.
The complete model with the time-varying force of infection is given by
the system of ordinary differential equations:
\begin{align}
   \label{eq:model}
 \frac{dS}{dt} &= -\beta(t)SI \\     \nonumber
 \frac{dI}{dt} &= \beta(t)SI - \tau I&\\      \nonumber
 \frac{dR}{dt} &= \tau I&
\end{align}  
where $S + I + R = 1 \: \forall\: t$. % \in [T_0, T_1]$. 
Of course, this is a rather simplified model, in which, for instance, the 
vector is omitted.
The rationale for this simplification is based on the ability of the 
empirically derived $R_t$  to incorporate the effects of the fluctuating vector 
populations. Although demography is not included in the model -- population is 
assumed to remain constant throughout an epidemic -- the population variation 
from year to year is taken into consideration as the prevalences are calculated 
for each year by dividing the number of cases by the official population 
estimated by the census. In Brazil, dengue affects all age groups which still 
haven't been exposed to all 4 serotypes. This is in contrast with countries in 
which the four serotypes have been endemic for a very long period of time, 
where dengue mostly affect the youth.

Also, although there are multiple circulating serotypes, our approach
can not discriminate between them due to the lack of serotype-specific data.
Nevertheless, this modelling strategy can still provide some insight into the 
disease dynamics and allows us to estimate the initial fraction of susceptibles 
$S_0$, a key epidemiological parameter.

@@

We take a Bayesian approach to the estimation of $S_0$.
First the incidence time series was divided into $J=13$ 
epidemic windows that corresponded to significant raises in incidence and 
normalized to lie on the $[0,1]$ interval.
For a given interval $j = \{ t_j^{\text{start}}, t_j^{\text{end}} \} $ we 
observe an incidence time series $\mathbf{Y_{j}}$.
We are thus interested in the posterior distribution
\begin{equation}
 \label{eq:S0post}
 p(S_{0j}|\mathbf{Y_{j}}) \propto l(\mathbf{Y_{j}}|S_{0j}, R_t, \tau 
)\pi(S_{0j}) 
\end{equation}
The likelihood $l(\mathbf{Y_{j}}|\cdot)$ is a approximated by a Normal 
distribution with fixed variance $\sigma^2$. This approximation is a numerical 
convenience since it confers better stability to the MCMC sampling.
In this estimation procedure we kept $R_t$ fixed at fixed at the posterior mean 
obtained as described above and fixed $\tau = 1/7\: \text{days}^{-1}$.
To complete the inference, we need to specify prior distributions for the 
parameters of interest.
We place a flat $\text{Beta}(1, 1)$ prior on $S_{0j}\:\forall j$

To approximate the posterior in~(\ref{eq:S0post}) we use Markov chain Monte 
Carlo techniques implemented in the Bayesian inference with Python 
(BIP)~\citep{pone2011} available at  
\url{http://code.google.com/p/bayesian-inference/}.
BIP uses a Differential Evolution Adaptive Metropolis (DREAM)~\citep{vrugt2008} 
scheme that efficiently samples from high-dimensional joint distributions using 
multiple adaptive chains running in parallel with delayed rejection (see 
\textbf{Appendix} for details).
Also, as the numerical integration routine implemented within BIP needs 
$\beta(t)$ to be available at 
arbitrary values of $t$, i.e., as continuous function of time because of the 
variable step size, we used linear interpolation to obtain values of $R_t$ for 
any time point.
In this study we used one chain per parameter, i.e, 3 chains for each run.
The chains were run until $5000$ samples were obtained after discarding $500$ 
burn-in samples.
Convergence of the parallel chains was verified at every 100 iterations by the 
calculation of the Gelman-Rubin's R (potential scale 
reduction factor), which approaches $1$ at convergence~\citep{brooks1998}.

@@

In this paper we propose a method to bypass the lack of serotype-specific case 
data by informing the time-varying force of infection with the instantaneous 
reproductive number, \rr which we calculate from aggregated data.
The main contribution of this paper can be summarized in the following items: 
(i) we develop a method to quantify uncertainty 
about \rr that is readily applicable to other diseases and; (ii) \rr is used to 
inform a dynamic epidemic model with time-varying force of infection 
in order to gain insight into the attack ratio of each epidemic; (iii) we 
propose an estimation procedure for circulating serotype's $S_0$ from 
aggregate case data, which is robust to epidemic sizes; (iv) AR estimates are 
provided for 18 years of Dengue epidemics in Rio de Janeiro, Brazil.

Figure~\ref{fig:rtseries} shows the $R_t$ series, according to 
Equation~\ref{eq:Rtestimate}~\citep{nishiura} along with the confidence bands 
derived 
in this paper. 
It can be seen that the inter-epidemic periods are characterized by $R_t$ being 
indistinguishable from 1.
Due to the intrinsic variability of the~\rr series, the examination of its 
credible intervals is essential to identify periods of sustained transmission.
The wider intervals between epidemics are due to the scarcity of cases during 
these periods. 
credibility intervals, and therefore offers protection against false alarms 
(see 
the section on tail behaviour in the \textbf{Appendix} for a detailed 
explanation).
\begin{center}
\begin{figure}[!h]
 \centering
 \includegraphics[width=16cm]{./plots/rt_series.png}
 \caption{{\bf Estimated time-series for $R_t$, along with 95\% credible 
intervals.} Top panel show reported cases from which $R_t$ is estimated.
As expected, uncertainty about \rr is greater when the case counts are low, for 
instance in the period $2003$--$2006$, which represented a big hiatus between 
major epidemics.
The intrinsic variability of \rr can be used to inform the time-varying force 
of infection, since it reflects variation in the vector population and other 
environmental factors such as temperature and seasonal variation.}
\label{fig:rtseries}
\end{figure}
\end{center}

A key epidemiological quantity is the attack ratio (AR) of an epidemic, a 
measure of morbidity and speed of spread which can be used to predict epidemic 
size and help efficient Public Health planning.
The AR depends fundamentally on the population at risk, which in the case of 
dengue is every naive (to a particular serotype) individual in the population.
Estimating the initial susceptible fraction $S_0$ for each epidemic is thus 
central to the estimation of the AR.
Methods for estimating the  number of susceptibles have been proposed 
before, for other diseases~\citep{bjornstad_dynamics_2002, 
wallinga_reconstruction_2003}.
These methods attempt to reconstruct the entire 
series of infectious and susceptibles for measles 
outbreaks from case data.
In the case of dengue, the full (multi-year/multi-epidemic)
series of susceptibles to all possible serotypes, cannot be reconstructed based 
solely on a deterministic transmission model, since the arrival/re-emergence of 
new serotypes (which are a stochastic events) can change drastically the pool 
of susceptibles throwing off any sequential estimation based on the incidence
dynamics.

Since there is very limited information regarding the actual proportions of 
each virus in circulation and most information available is about the 
predominant serotypes for some epidemics in the period of 
study only~\citep{macedo_virological_2013}, we propose the use of a 
simplified a single strain model.
The main argument we put forward is that by conditioning on the \rr series, we 
implicitly take into account the variability introduced by the 
co-circulation of multiple serotypes and heterogeneous levels of immunity in 
the general population.
We sought to deal with all important sources of uncertainty impinging on the 
estimation of the AR of a dengue epidemic, but not all could be satisfactorily 
addressed in this analysis.
For instance, in any given epidemic there is a large number of 
mild and asymptomatic cases, which nevertheless acquire immunity.
It is estimated that for every case reported, up to 10-20 are not seen by 
health 
authorities~\citep{luz_disability_2009}.
Another source of uncertainty is under-reporting of diagnosed cases, which is a 
serious issue in the health care systems of many developing countries such as 
Brazil. 
Duarte and Fran\c{c}a (2006)~\citep{duarte_data_2006}, estimated 
the sensitivity of Dengue reporting for hospitalized patients in 
Belo-Horizonte, Brazil to be of 63\%, meaning that approximately 37\% of the 
suspected Dengue cases go unreported.  
Lastly, demography and migrations affect the number of susceptible in ways 
which are not easy to fully determine.

@@

Table~\ref{tab:AR} contains the attack ratios and medians of the $S_0$ 
estimated for each epidemic/outbreak.
 Underreporting of cases, which is known to exist but of which exact figures 
cannot be determined, will lead to underestimation $S_0$.
However, the attack ratio shall remain unbiased as the underreporting affects 
both the numerator and the denominator of equation~\ref{eq:AR2}.
It is interesting to notice that the larger epidemics, in terms of peak size 
are not the one with the greater attack ratios.
This stresses the importance of knowing the immunological structure of the 
population.
Knowing the $S_0$ for the circulating viruses we can order to more 
accurately assess the potential impact of a coming epidemic, since particularly 
virulent types, can be rendered less of a threat by a low $S_0$.
Hon\'orio et al.\citep{honorio2009spatial} conducted a serological survey in 
three separate localities within the city, right before the 2008 epidemic, the 
authors report seroprevalences varying from 56-77.4\% which is compatible with 
our prediction of 87.5\% ($1-S_{0,2007}$) for the entire city, considering that 
we underestimate $S_0$ due to the underreporting of cases.
